Abstract Spatial Specification of Stencil Computations

### 2016-07-03 paper.tex
Mathematical model nearly finished

\newcommand{\stenFwd}[2]{\term{forward} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwd}[2]{\term{backward} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCen}[2]{\term{centered} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
  \term{dim=}#2)}

@@

\newcommand{\singleEntry}[2]{\textbf{J}_{#2}^{#1}}

@@

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{specTop} \;
                        \texttt{::} \; v
  \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[1em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{specTop} ::= & [\nonterm{approxMod},] \; \nonterm{spec} \\
\nonterm{spec} ::= & [\nonterm{mod}] \; [\nonterm{region}] \\
\nonterm{mod} ::= & [\nonterm{modR},] \; [\nonterm{modRefl},] \; [\nonterm{modIrefl},] \\
\nonterm{modR} ::= &
 \term{readOnce} \\
\nonterm{modRefl} ::= & \term{reflexive} \, (\term{dims=}\dimTy \{\term{,} \; \dimTy\}) \\
\nonterm{modIrrefl} ::= & \term{irreflexive} \, (\term{dims=}\dimTy \{\term{,}
          \dimTy\}) \\
\nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \stenFwd{\mathbb{N}_{>0}}{\dimTy}  \\
\mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
\mid \; & \nonterm{region} \; \term{*} \; \nonterm{region} \\
\mid \; & \nonterm{rvar}  \\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

@@

To save space we use abbreviations
\term{fwd}, \term{bwd}, and \term{cen} for \term{forward},
\term{backward}, and \term{centered} regions respectively.  The
equations reveal something of their semantics.  The first six
equations (with labels of the form $\ldots \texttt{+} \ldots$)
explain the behaviour of overlapping regions with sum
\term{+}. The next six equations explain the algebraic behaviour of
\term{+} and \term{*}.  Together, we see that \term{+} is an
associative, commutative, idempotent binary operator that distributes
with \term{*}, which is associative and commutative. This distribution
of \term{*} over \term{+} is key as it is used internally to give a
normal form for stencil specifications akin to Disjunctive Normal Form
(DNF) (where \term{+} is taken as ``disjunction'' and \term{*} as the
``conjunction''). We revisit this normalisation in Section~\ref{}.

@@

Figure~\ref{fig:model} outlines the core rules for regions
via the interpretation function $\interp{-}$.  We explain the
intermediate notations and definitions:

\begin{definition}A \emph{single-entry vector} of size $n$, denoted
$\singleEntry{r}{n}$, is a vector
where the $r^{th}$ entry is $1$ and all others are $0$, \eg{},
$\singleEntry{1}{2} = \vtwoh{1}{0}$.
\end{definition}

\begin{definition}The product of two models $\otimes$ is defined:
%%
\begin{align*}
M \otimes N & =
  \{(u \bowtie v)_i \mid i \in \{1, \ldots, 2^n\}, (u, v) \in M \times N\}
\end{align*}
The $\bowtie$ operation, which we call \emph{pairwise permutation}
(described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
$n$-vectors). Using this, the $\otimes$ operation constructs the set of every
pairwise permutation for every pair $(u, v)$ of $n$-vectors in the Cartesian
product of $M \times N$.

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-determinstically picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{ccc}
0 & 1 & 2 \\
0 & 1 & 5 \\
0 & 4 & 2 \\
0 & 4 & 5 \\
3 & 1 & 2 \\
3 & 1 & 5 \\
3 & 4 & 2 \\
3 & 4 & 5
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
defined the pairwise pemutation matrix as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$. Recall that the vectors are defined over the domain
$\mathbb{Z} \cup \{\infty\}$.
For addition and multiplication, we take
 $\infty$ to be an absorbing element. Thus, if either
vector in a pairwise permutation contains $\infty$ in
the $j^{th}$ entry, then $j^{th}$ entry of every pairwise
permutation is $\infty$, \eg{},
%%
\begin{equation*}
\vtwoh{0}{1} \bowtie \vtwoh{2}{\infty} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{cc}
0 & \infty \\
0 & \infty \\
2 & \infty \\
2 & \infty
\end{array}
\right]
\end{equation*}

@@

\begin{figure}
\input{model.tex}
%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
%  the multi-set $A$}
\caption{Set model of specifications}
\label{fig:model}
\end{figure}

>>>>

\newcommand{\stenRefl}[1]{\term{reflexive} \, (\term{dim=}#1)}
\newcommand{\stenFwd}[2]{\term{forward} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwd}[2]{\term{backward} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCen}[2]{\term{centered} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenReflS}[1]{\term{refl} \, (\term{dim=}#1)}
\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
  \term{dim=}#2)}

@@

\newcommand{\singleEntry}[2]{\textbf{J}_{#2}^{#1}}
\newcommand{\zeroEntry}[2]{\textbf{K}_{#2}^{#1}}

@@

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{specTop} \;
                        \texttt{::} \; v
  \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[1em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{specTop} ::= & [\nonterm{approxMod},] \; \nonterm{spec} \\
\nonterm{spec} ::= & [\nonterm{modMult},] \; \nonterm{refRegion} \\
\nonterm{refRegion} ::= & [\nonterm{modRefl},] \; \nonterm{region}\\
\nonterm{modMult} ::= & \term{readOnce} \\
\nonterm{modRefl} ::= & \term{irreflexive} \, (\term{dims=}\dimTy \{\term{,}
          \dimTy\}) \\
\nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::=
        & \stenFwd{\mathbb{N}_{>0}}{\dimTy}  \\
\mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \stenRefl{\dimTy}  \\
\mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
\mid \; & \nonterm{region} \; \term{*} \; \nonterm{region} \\
\mid \; & \nonterm{rvar}  \\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

@@

To save space we use abbreviations
\term{refl}, \term{fwd}, \term{bwd}, and \term{cen} for \term{reflexive},
\term{forward}, \term{backward}, and \term{centered} regions respectively.
The equations reveal something of their semantics.  The first six
equations (with labels of the form $\ldots \texttt{+} \ldots$)
explain the behaviour of overlapping regions with sum
\term{+}. The next six equations explain the algebraic behaviour of
\term{+} and \term{*}.  Together, we see that \term{+} is an
associative, commutative, idempotent binary operator that distributes
with \term{*}, which is associative and commutative. This distribution
of \term{*} over \term{+} is key as it is used internally to give a
normal form for stencil specifications akin to Disjunctive Normal Form
(DNF) (where \term{+} is taken as ``disjunction'' and \term{*} as the
``conjunction''). We revisit this normalisation in Section~\ref{}.

@@

Figure~\ref{fig:region-model} outlines the core rules for regions
via the interpretation function $\interp{-}$.  We explain the
intermediate notations and definitions:

\begin{definition}A \emph{single-entry vector} of size $n$, denoted
$\singleEntry{r}{n}$, is a vector
where the $r^{th}$ entry is $1$ and all others are $\infty$, \eg{},
$\singleEntry{1}{2} = \vtwoh{1}{\infty}$.
\end{definition}

\begin{definition}A \emph{zero-entry vector} of size $n$, denoted
$\zeroEntry{r}{n}$, is a vector where the $r^{th}$ entry is $0$ and all others
are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
\end{definition}

\begin{definition}The product of two models $\otimes$ is defined:
%%
\begin{alignat*}{2}
& dims(M_m,N_n) =
  \bigg\{ x \; \bigg| \;
    \parbox{4.1cm}{
      $x \in \{1, \ldots, m\} \wedge M_x \neq \infty \; \vee \;$ \\
      $x \in \{1, \ldots, n\} \wedge N_x \neq \infty$
                  } \bigg\} \\
& M_n \otimes N_n =
  \Bigg\{x \; \Bigg| \;
    \parbox{4.8cm}{$i \in \{1, \ldots, 2^n\},$
                   $(u, v) \in M_n \times N_n,$ \\
                   $x = (u \bowtie v)_i,$ \\
                   $\forall j \in dims(M_n,N_n) (x_j \neq \infty) $
                  } \Bigg\}
\end{alignat*}
The $\bowtie$ operation, which we call \emph{pairwise permutation}
(described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
$n$-vectors). Using this, the $\otimes$ operation constructs the set of every
pairwise permutation for every pair $(u, v)$ of $n$-vectors in the Cartesian
product of $M \times N$.

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-determinstically picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{ccc}
0 & 1 & 2 \\
0 & 1 & 5 \\
0 & 4 & 2 \\
0 & 4 & 5 \\
3 & 1 & 2 \\
3 & 1 & 5 \\
3 & 4 & 2 \\
3 & 4 & 5
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
defined the pairwise pemutation matrix as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$. Recall that the vectors are defined over the domain $\mathbb{Z} \cup
\{\infty\}$. In pairwise permutation $\infty$

@@

\begin{figure}
\input{region-model.tex}
%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
%  the multi-set $A$}
\caption{Set model of regions}
\label{fig:region-model}
\end{figure}

\begin{figure}
\input{spatial-model.tex}
\caption{Model of specifications}
\label{fig:spatial-model}
\end{figure}

\begin{figure}
\input{consistency.tex}
\caption{Consistency function for a model and a set of indices}
\label{fig:spatial-model}
\end{figure}
